A forest is more than its trees: haplotypes and inferred ARGs

### 2024-10-28 paper/writeup.tex
.

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\similarity}{\operatorname{sim}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}

@@

\comment{
    Idea: look at statistical properties of unary regions:
    look at how much 'unary regions' there is in simulation
    and how much we can put back in,
    as a function of time.
}

@@

\subsection{Tree sequence compression and computation}

@@

One might expect that a reduction in edges would also lead to a reduction in computation time
(for instance, of the class of statistics described in \citet{ralph2020efficiently}.
This turns out not to be the case, as shown in \figref{fig:num_edges}.
The reason for this is clear if we
return to the algorithms in \citet{ralph2020efficiently}:
each edge added or removed requires an update
to the state of the parent node \emph{and every node above it in the tree}.
So, even if an edge extends for a long segment of genome,
something has to happen to it every time the subtree below it changes.
Extending haplotypes does not change at all the locations in the tree affected by subtree changes.
\comment{If the increase in computation time holds up:}
In fact, introducing unary nodes to branches
\emph{increases} the distance from a typical node up to the root,
and so increases the work required (and also computation time, as seen in \figref{fig:num_edges}).

\subsection{Reconstructing haplotypes of true trees}

Our first task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
are indeed correct.
To do this, we simulate ARGs containing full haplotypes using \msprime
as described in the Methods.

@@

\subsection{Extending haplotypes in inferred trees}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		\caption{Left: Dissimilarity \newline Right: True Proportion }
	\end{center}
	\label{fig:discrepancy-over-sample}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
		\caption{Left: Dissimilarity \newline Right: True Proportion}
	\end{center}
	\label{fig:discrepancy-over-seqlen}
\end{figure}

@@

@@

\begin{center}
\begin{table}[hbt]
\begin{tabular}{|c|c|c|c|c|}
	\hline
	Initial & Iteration 1 & Iteration 2 & Iteration 3 & Iteration 4 \\
	\hline
	\hline
	113463	&	76301	&	76059	&	76057	&	76057 \\
	\hline
	113266 & 76084	&	75835	&	75833	&	75833 \\
	\hline
	114489 & 76703 & 76436 & 76434 & 76434 \\
	\hline
	114086	& 76550	& 76294	& 76291	& 76291\\
	\hline
\end{tabular}
\caption{Number of edges for each iteration of \textit{extend haplotypes} applied to 
simulated tree sequences with sample size $10^4$ and length $5\times 10^7$. 
Each of the simulations terminated by the fourth iteration, however 99\% of the edges
are removed within the first iteration.}
\label{tab:edge-counts}
\end{table}
\end{center}

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\similarity}{\operatorname{sim}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}

@@

% Outline:
% 1. benchmarking: edge reduction and runtime
% 2. correctness:

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tree sequence compression and computation}

@@

This reduction in edges can also lead to a reduction in computation time.
Indeed, \figref{fig:num_edges} shows that computation time is reduced by 10--20\%
for a typical statistic (here, Tajima's $D$),
computed in an efficient iterative manner along the genome as implemented in \tskit.
This is actually somewhat surprising:
as described in \citet{ralph2020efficiently},
each edge added or removed requires an update
to the state of the parent node \emph{and every node above it in the tree}.
So, even if an edge extends for a long segment of genome,
something has to happen to it every time the subtree below it changes,
which is not affected by extending haplotypes.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Accuracy with true trees}

Our next task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
are indeed correct -- i.e., that in addition to compression, we are also gaining information.
To do this, we simulate ARGs containing full haplotypes using \msprime{}
(as described in the Methods).

@@

%%%%%%%%%%%%%%%%%%
\subsection{Inferred ARGs}

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes;
to what degree are they taking advantage of this information already?
Here, we study the behavior of \tsinfer{},
leaving a comparison of different ARG inference methods for future work.
To study this, we simulated ARGs containing unary-spanning haplotypes (as above),
then performed various operations to them,
reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:discrepancy}.

The first thing we see is that coalescent nodes are unary
over a substantial portion of their spans:
since the \texttt{simplify} operations removes these unary spans,
the ``simplified'' line on the right of Figure~\ref{fig:discrepancy} shows
the proportion of nodes' spans on which they are not unary.
For instance, on an ARG of 10 samples on a sequence of length $5 \times 10^7$,
only about 15\% of the total spans are retained after simplification;
this rises to around 50\% for 1,000 samples (top right plot)

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Left:} Discrepancy to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (equation~\eqref{eq:discrepancy}).
        \textbf{Right:} Agreement to the true ARG,
        as proportion of the true ARG that is represented.
        The largest number of trees in the bottom row corresponds
        to the sequence length in the top row;
        see text for more details.
    }
    \label{fig:discrepancy}
\end{figure}

@@

\paragraph{Inference methods}
\comment{Possibly an overview of what the different inference methods do with unary spans?}

@@

\begin{table}[hbt]
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
	\hline
	Initial & Iteration 1 & Iteration 2 & Iteration 3 & Iteration 4 \\
	\hline
	\hline
	113463	&	76301	&	76059	&	76057	&	76057 \\
	\hline
	113266 & 76084	&	75835	&	75833	&	75833 \\
	\hline
	114489 & 76703 & 76436 & 76434 & 76434 \\
	\hline
	114086	& 76550	& 76294	& 76291	& 76291\\
	\hline
\end{tabular}
\caption{Number of edges for each iteration of \textit{extend haplotypes} applied to 
simulated tree sequences with sample size $10^4$ and length $5\times 10^7$. 
Each of the simulations terminated by the fourth iteration, however 99\% of the edges
are removed within the first iteration.}
\label{tab:edge-counts}
\end{center}
\end{table}
